---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


1. Loading and preprocessing the data
  1.1 Reading the data into R

```{r, echo=TRUE}
activity <- read.csv("activity.csv", colClasses = c("numeric", "Date", "numeric"))
```
  1.2 Adding the weekend / weekday indicator necessary later on
  
```{r, echo=TRUE}
activity$day_of_week <-weekdays(activity$date)
activity$isWeekend <-ifelse(activity$day_of_week == "Sunday" | activity$day_of_week == "Saturday", 1, 0)

```

2. What is mean total number of steps taken per day?
  2.1 Calculate the total number of steps taken per day 

```{r, echo = TRUE}

require(plyr)
activityByDay <-ddply(activity, ~date, summarise, sum=sum(steps,na.rm = TRUE ), mean=mean(steps, na.rm = TRUE), median=median(steps, na.rm = TRUE) )
```
  2.2 Make a histogram of the total number of steps taken each day
```{r, echo = TRUE}
plot(activityByDay$date, activityByDay$sum,  type = "h", lwd=15, lend=1, main = "total number of steps taken each day", xlab = "day", ylab = "steps" )

```

  2.3 Calculate and report the mean and median of the total number of steps taken per day (table suppressed because it is +200 lines)
```{r, eval=FALSE}
c(activityByDay$date, activityByDay$mean)
c(activityByDay$date, activityByDay$median)

```


3. What is the average daily activity pattern?

  3.1 Make a time series plot
```{r, echo = TRUE}
activityByInt <-ddply(activity, ~interval, summarise, mean=mean(steps, na.rm = TRUE) )

plot(activityByInt$interval, activityByInt$mean, type = "l", col="blue", main = "Mean activity over the day", xlab = "interval", ylab = "steps")
```

  3.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
indexMax <- which.max(activityByInt$mean)
print(activityByInt$mean[indexMax])

```


4. Imputing missing values
  4.1 Calculate and report the total number of missing values in the dataset
```{r, echo = TRUE}
sum(is.na(activity$steps))
```

  4.2 Devise a strategy for filling in all of the missing values in the dataset - Fill all NAs with 0 (otherwise overreporting)
  4.3 Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r, echo = TRUE}
activityNoNA <- activity
activityNoNA$steps[is.na(activityNoNA$steps)] <- 0

activityNoNAByDay <-ddply(activityNoNA, ~date, summarise, sum=sum(steps ), mean=mean(steps), median=median(steps) )

```
  4.4 Make a histogram of the total number of steps taken each day 

```{r, echo = TRUE}
plot(activityNoNAByDay$date, activityNoNAByDay$sum,  type = "h", lwd=15, lend=1 , main = "Total steps taken each day", xlab = "day", ylab = "steps")
```
  4.5 Calculate and report the mean and median total number of steps taken per day
```{r, eval = TRUE}
print(activityNoNAByDay$mean)
print(activityNoNAByDay$median)

```
  4.6 Do these values differ from the estimates from the first part of the assignment? (No) What is the impact of imputing missing data on the estimates of the total daily number of steps? (Replacing NA with 0 does not impact mean & median)
```{r, echo = TRUE}
sum(activityNoNAByDay$mean != activityByDay$mean, na.rm = TRUE)
sum(activityNoNAByDay$median != activityByDay$median, na.rm = TRUE)

```

5. Are there differences in activity patterns between weekdays and weekends? (Yes)

```{r, echo = TRUE}

activityWeekday <-ddply(activity[activity$isWeekend == 0,], ~interval, summarise, mean=mean(steps, na.rm = TRUE) )
activityWeekend <-ddply(activity[activity$isWeekend == 1,], ~interval, summarise, mean=mean(steps, na.rm = TRUE) )
par(mfrow=c(2,1))

plot(activityWeekday$interval, activityWeekday$mean, type = "l", col = "blue", main = "Weekday", xlab=("interval"), ylab=("Mean Steps"))
plot(activityWeekend$interval, activityWeekend$mean, type = "l", col = "red" , main = "Weekend", xlab=("interval"), ylab=("Mean Steps"))

```
